---
title: "ESM 244 Lab 1"
author: "Narjes Mathlouthi"
date: "1/8/2022"
output: html_document
---

```{r setup, include=TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(here)
library(sf)
library(tmap)

##install.packages("tmap")
##update.packages(ask = FALSE)

```


### Read in the data

```{r}
sf_trees<- read_csv(here("data", "sf_trees", "sf_trees.csv"), show_col_types = FALSE)

```


### Part 1: Wrangling and ggplot review
**Example 1:** Find counts of observation by `legal_status` & wrangle a bit.

```{r}

### method 1 : group_by() %>%  summarize()

sf_trees %>% 
  group_by(legal_status) %>% 
  summarize(tree_count = n())

### method 2: different way plus a few new functions

top_5_status <- sf_trees %>% 
  count(legal_status) %>% 
  drop_na(legal_status) %>% 
  rename(tree_count = n) %>% 
  relocate (tree_count) %>% 
  slice_max(tree_count, n = 5) %>% 
  arrange(desc(tree_count))
  

```
Make a graph of the top 5 from above


```{r}
ggplot(data = top_5_status, aes(x = fct_reorder(legal_status, tree_count), y = tree_count)) + 
  geom_col(fill= "darkgreen")+
  labs(x="legal status", y = "Tree count") +
  coord_flip()+
  theme_minimal()
```

**Example 2:** Only keep observation where legal status is "Permitted Site" and caretaker is "MTA", and store as `permitted_data_df`

shift-cmd-c to comment/uncomment quickly 

```{r}
# sf_tree$legal_status %>%  unique()
# unique(sf_trees$caretaker)

permitted_data_df <- sf_trees %>% 
  filter(legal_status == 'Permitted Site' & caretaker == "MTA")
```


**Example 3:** Only keep Blackwood Acacia trees, and then only keep columns `legal_status`, `date`, `longitude` and as `blackwood_accacia_df`


```{r}
blackwood_acacia_df <- sf_trees %>% 
  filter(str_detect(species, 'Blackwood Acacia')) %>% 
  select(legal_status,date, lat= latitude, lon = longitude)

### Make a little graph of locations 

ggplot(data = blackwood_acacia_df, aes(x = lon, y = lat)) +
  geom_point(color = 'darkgreen')
  
```

**Example 4:** use `tidyr::seperate`

```{r}

sf_trees_sep <- sf_trees %>% 
  separate(species, into = c('spp_scientific', 'spp_common'), sep = ' :: ')
```

**Example 5 :** use `tidyr::unite()`

```{r}
ex_5 <- sf_trees %>% 
  unite('id_status', tree_id, legal_status, sep = '_COOL_' )
```





